Follow-up Samples

### 2017-04-11 submission/manuscript.Rmd

***The Bacterial Community:***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome in patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples. We found that carcinoma patients had a more dissimilar bacterial community between these samples than those with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001) [Figure 1A]. Although no significant differences were observed between advanced adenoma and carcinoma there was an increase in the dissimilarity of pre- and post-treatment samples from adenoma (`r paste(format(change_summary["thetayc", "adn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "adn_SD"], digits = 2))` (mean `r paste(label = "\u00B1")` SD)) to advanced adenoma (`r paste(format(change_summary["thetayc", "srn_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "srn_SD"], digits = 2))`) to carcinoma (`r paste(format(change_summary["thetayc", "crc_mean"], digits = 2), label = "\u00B1", format(change_summary["thetayc", "crc_SD"], digits = 2))`) [Figure 1A]. The bacterial community structure pre- post-treatment was visualized using NMDS for adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary["Adenoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), advanced adenoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary["SRN", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05), and carcinoma [Figure 1D] (PERMANOVA `r ifelse(beta_summary["Carcinoma", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05). Interestingly, when pre- and post-treatment samples were compared, regardless of whether they were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary["All", "PERMANOVA"] < 0.05, invisible("<"), invisible(">"))` 0.05). There was no difference between pre- and post-treatment samples when investigating alpha diversity metrics for adenoma, advanced adenoma, or carcinoma [Table S1]. Additionally, there was also no difference in the relative abundance of any specific OTU between these samples for adenoma, advanced adenoma, or carcinoma only [Figure S1].

***Adenoma Model:***
Using normal and adenoma individuals from a separate cohort we created an adenoma model to help assess whether those with adenoma in this cohort changed towards normal. The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(adn_general_summary["full", "AUC"], digits = 2)` [Figure S2A]. There was a total of `r length(rownames(adn_imp_vars))` OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "adn" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05). There was also no significant decrease in the positive probability of adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2A].

***Advanced Adenoma Model:***
Using normal and advanced adenoma individuals from a separate cohort we created an advanced adenoma model to help assess whether those with advanced adenoma in this cohort changed towards normal. The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification having an AUC of `r format(srn_general_summary["full", "AUC"], digits = 2)` [Figure S2B]. There was a total of `r length(rownames(srn_imp_vars))` OTUs in the advanced adenoma model [Figure S3B]. Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "SRN" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05) and no significant decrease in the positive probability of advanced adenoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["SRN", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2B].

***Carcinoma Model:***
Using normal and carcinoma individuals from a separate cohort we created a carcinoma model to help assess whether those with carcinoma in this cohort changed towards normal. The range of model AUC's from 100 runs of 20 repeated 10-fold cross-validation was `r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)` with the AUC of the model used for classification being `r format(crc_general_summary["full", "AUC"], digits = 2)` [Figure S2C]. Interestingly, the AUCs improved from adenoma to advanced adenoma to carcinoma [Figure 2]. There was a total of `r length(rownames(crc_imp_vars))` OTUs in the carcinoma model [Figure S3C]. Similar to the adenoma and advanced adenoma models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* were also important for carcinoma classification [Figure S3C]. Also, like the adenoma and advanced adenoma models there was a significant difference between the actual and predicted disease classification (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "crc" & comparison == "McnemarPValue")[, "all"] < 0.05, invisible("<"), invisible(">"))` 0.05). There was a significant decrease in the positive probability of carcinoma between pre- and post-treatment samples (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or advanced adenoma. The one individual still positive for carcinoma after treatment had an increase in carcinoma positive proability between their pre- and post-treatment sample [Figure 2C].

***Common OTUs:***
Next, to identify which OTUs were important at all stages of disease, we identified common predictive OTUs within the adenoma, advanced adenoma, and carcinoma models. When we compared the three different models with each other there were a total of `r length(rownames(common_vars_summary))` common OTUs. Some of the most common taxonomic identifications belonged to *Bacteroides*, *Blautia*, *Anaerostipes*, *Lachnospiraceae*, and *Dorea*. The vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2]. The rank of these OTUs, based on median mean decrease in accuracy (MDA), varied higly depending on whether they were in the adenoma, advanced adenoma, or carcinoma model [Figure 3]. As an example, *Bacteroides* (OTU85) was ranked `r filter(crc_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ in the carcinoma model but in both the adenoma and advanced adenoma model ranked `r filter(adn_varsMDA, otu == "Otu000085") %>% select(rank)`^th^ and `r filter(srn_varsMDA, otu == "Otu000085") %>% select(rank)`^st^ respectively [Figure 3]. .

***Treatment Affects on Community:***
After observing these changes from treatment we assessed the possible confounding impact of chemotherapy or radiation, in the carcinoma group, and surgical resection, in the adenoma group, on the observed results. In the carcinoma group neither chemotherapy nor radiation provided any additional change from initial sample over surgery alone (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S3]. For the adenoma group there was a single difference in observed OTUs (sobs) between those that received surgerical resection and those that had regular polyp removal (P-value `r ifelse(min(comb_adn_surg_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05) [Table S4]. For the surgical resection comparison, adenoma and advanced adenoma were combined due to the low number of resection occurances in these two groups. Using a fisher exact test there was no difference in the proportion of those receiving surgerical resection between the adenoma and advanced adenoma groups (P-value `r ifelse(comb_adn_surg_summary["prop_surg_adn_srn", "pvalue"] < 0.05, invisible("<"), invisible(">"))` 0.05). This data suggests that the microbiome changes observed in the carcinoma group were mostly a result from surgerical resection and not from chemotherapy or radiation.

>>>>

***Treatment alters the bacterial community structure of patients diagnosed with colonic lesions.***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiome of patients with adenoma (N=`r change_summary["thetayc", "adn_n"]`), advanced adenoma (N=`r change_summary["thetayc", "srn_n"]`), or carcinoma (N=`r change_summary["thetayc", "crc_n"]`) had any broad differences between pre- and post-treatment samples (Table 1). The structure of the microbial communities of the pre- and post-treatment samples differed [Figure 1A]. We found that the communities obtained pre- and post-treament among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P>0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1BCD]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value`r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value`r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value`r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value>0.05) [Table S1].

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups [Figure S1]. It is likely that high inter-individual variation and the patchy distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subject in each group, we restricted our models to only incorporate 10 OTUs. Despite this restriction, the models performed well (adenoma AUC=, advanced adenoma AUC=, carcinoma AUC= ). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

***Post-treatment samples from patients with carcinoma share***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of 423 individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma (Table 2). We then constructed Random Forest models to classify the pre and post-treatment samples as having their original diagnosis or having a normal colon. The models performed well (adenoma AUC=`r format(min(adn_AUC_run_summary$ROC), digits = 2)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2)`, advanced adenoma AUC=`r format(min(srn_AUC_run_summary$ROC), digits = 2)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2)`, carcinoma AUC=`r format(min(crc_AUC_run_summary$ROC), digits = 2)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2)`; Figure S2). The OTUs that were most important in the models for differentiating between patients with normal colons and those with adenomas or advanced adenomas were dominated by populations commonly considered as commensals (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, etc. [Figure?]). Although many of these OTUs were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [Figure ?]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the colon more closely resembled that of a patient with a normal colon. There was not a significant change in the positive probability for the adenoma or advanced adenoma groups [Figure 2]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased. In fact, only 3 of the 26 patients (12%) that were initially diagnosed with a carcinoma had a higher positive probability after treatment. One of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that although there are changes in the microbiota associated with treatment, those experienced by the patients in the carcinoma group were the only ones where the change was directed towards what would be found in a normal colon

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=4; advanced adenoma, N=4) or polyp removal during colonoscopy (adenoma, N=4; advanced adenoma, N=4) and those with carcinomas received surgical resection (N=~14), surgical resection with chemotherapy (N=~6), and surgical resection with chemotherapy and radiation (N=~6). We focused on the patients with carcinoma and pooled those patients that received chemotherapy with those that received chemotherapy and radiation to improve our statistical power. We did not observe a significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P<0.XXX). Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P=0.XX). Finally, the change in the positive probability was not significantly different between the two treatment groups (P=0.XX). Due to the relatively small number of samples in each treatment group, it is difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota
